---
#title: "MATH2349 Semester 1, 2018 - ADD DESCRIPTIVE TITLE"
#author: "Alistair Grevis-James (s3644119) and Chris Kaias (sXXXXXX)"
#subtitle: Assignment 3
#output:
#  html_notebook:
#      toc: true
#      toc_float: true
---

<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
options(knitr.table.format = "html")
```

### Executive Summary

Data blah blah and ... These data were imported as `xxx` and `yyy`, respectively.

### Required packages 

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
#The following packages were used whilst completing the report (with annotation)
library(readr) # Used to import .csv files
library(dplyr) # Section Y
library(stringr) # Section Z
library(knitr) # Section X
library(tidyr)
library(lubridate)
#xlsx #readxl #foreign #gdata #rvest #dplyr  #deductive #validate #Hmisc #outliers
#MVN #infotheo #MASS #caret #MLR #ggplot #base R functions
```

### Importing the Data

The `readr` function was used to import the data as shown below. The data were imported as `gvd` for Gun Violence Data and `acs2015` for American (US) 2015 Census Data.

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
gvdZip <- "gun-violence-data.csv.zip"
outDir<-"/Users/alistairgj/Documents/GitHub" # Change output directory for unzipped .csv file as needed
unzip(gvdZip, exdir=outDir)
setwd("/Users/alistairgj/Documents/GitHub")
gvd <- read.csv("gun-violence-data.csv") # Importing Gun Violence Data
setwd("/Users/alistairgj/Documents/GitHub/MATH2349_Assignment3V2")
acs2015 <- read.csv("acs2015_county_data.csv") # Importing American (US) Census Data 2015
```

### Data Summary

Three data sets from two sources was used during this analysis.

Gun Violence Data a Comprehensive record of over 260k US gun violence incidents from 2013-2018 (italics), was compiled by James Ko and made available on kaggle.com (footnote). These data contain all recorded gun violence incidents in the US between January 2013 and March 2018, inclusive. These data were imported as `gvd`.

https://www.kaggle.com/jameslko/gun-violence-data/data = gun_violence_data

US Census Demographic Data Demographic and Economic Data for Tracts and Counties (italics), was collected by the US Census Bureau and made available on kaggle.com (footnote). These data contain all cencus data from the 2015 federal census. These data were split into two data sets, the first being county data, and the second being tract data (The "Census Tract" is an area roughly equivalent to a neighborhood established by the Bureau of Census for analyzing populations. They generally encompass a population between 2,500 to 8,000 people).

https://www.kaggle.com/zimeiyang/2015-us-census-demographic-data/data = acs_2015


### Initial Check of Gun Violence Data

The dataset Gun-Violence Data was imported as `gvd`. Using `str()` it was checked for dimensions (239677 observations of 29 attributes), column names and data types.

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
str(gvd)
```

### Filtering Gun Violence Data for values inclusive of the year 2015

The Gun Violence Dataset contains data from 2013 till 2015, while the American Census Data is from 2015. We will therefore create a data subset of the Gun Violence Data from the year 2015 only. As previously determined (`str(gvd)`), the `gvd$date` attribute is present as a factor with 1725 levels (1725 different date values). This attribute will be converted to date format (YYYY-MM-DD) using `as.Date`. This conversion will be checked with `str()`, below.

```{r}
gvd$date <- as.Date(gvd$date) # Converting the gvd$date column and applying to the gvd dataset
str(gvd$date) # Checking the conversion of gvd$date
```

The American (US) Census Data contains values from the year 2015 only, and as such we will subset the Gun Violence Data accrodingly. The first methodology explored for extracting YYYY = 2015 value from `gvd$date` 

Intially `gvd$date <- format.Date(gvd$date, format="%Y")` was used to convert YYYY-DD-MM to YYYY values. This was then filtered
#gvd$date_year <- filter(gvd, date_year == "2015") # Filtering Gun Violence Data for 2015 only

Want to retain all date information but in one column at the moment.
- Previous attempt 

```{r}
year2015_boolean <- as.logical(year(gvd$date) == "2015") #Creating 2015 TF
summary(year2015_boolean)
```

```{r}
gvd2015 <- gvd[year2015_boolean,]
```

```{r}
dim(gvd2015)
```

```{r}
sapply(gvd, class)
```

```{r}
gvd2015$date <- factor(gvd2015$date)
str(gvd2015$date) # Confirming only 2015 data is present
```

```{r}
sapply(gvd, class) # Checking the data types for 'gvd'
```

### Filering for only one State

The data frame `gvd2015` has dimensions of 53579 instances over 29 attributes, we will now check the values of `gvd2015$state` using the aggregate function.

```{r}
aggregate(city_or_county ~ state, gvd2015, function(x) length(unique(x)))
```



### Dropping Attributes from GVD

After the GVD dataframe was subset for 2015, the attributes were examined to determine their value in further analysis. The data frame `gvd2015` has dimensions of 53579 instances over 29 attributes. 

```{r}
# Insert display of attributes / explain, narrate attributes ...
gvd2015_check1 <- select(gvd2015, date, address, incident_url, source_url, incident_url_fields_missing)
#head(gvd2015_check1, n=6)
```

The following attributes will be dropped from GVD (2015); `date` as this is all for the year 2015 now, `address` as this level of fine detail is not required, `incident_url`, `source_url` & `incident_url_fields_missing` as these have no value for our analysis.
+ `gun_stolen` to sparse, `gun_type` to sparse, `location_description` to sparse, `notes` no value, `participant_age_group` we already have participant age,  `participant_name`  `participant_relationship`  `sources`     

```{r}
gvd2015_keep <- select(gvd2015, incident_id, state, city_or_county, n_killed, n_injured, congressional_district, incident_characteristics, latitude, longitude, n_guns_involved, participant_age, participant_age_group, participant_gender, participant_status, participant_type, state_house_district, state_senate_district)
```

```{r}
dim(gvd2015_keep)
```



```{r}
aggregate(state ~ state, gvd2015_keep, function(x) length(unique(x)))
```

For our analysis the State California will be chosen. *Talk about varied socio-economic status..?*

```{r}
gvd2015_keep_cali <- filter(gvd2015_keep, state == "California")
```

```{r}
dim(gvd2015_keep_cali)
```

#### Exploring the GVD 2015 California Data Subset

```{r}
colnames(gvd2015_keep_cali)
```

The participant age and gender attributes may contain ... (good data for analytics)

```{r}
gvd2015_keep_cali[1:4, c(4,5,7,10,11)]
```

```{r}
gvd2015_keep_cali[1:4, c(12:15)]
```

In both attibutes, multiple values are contained. 

*Not tidy, multiple values in one column*

```{r}
gvd2015_keep_cali$matches <- str_match_all(gvd2015_keep_cali$participant_age, "\\d+::(\\d+)")
ages = c(c(0,15)) # First age range is under 15s
for (i in seq(16, 76, 5)) # The other age ranges are every 5 years 
{  ages <- c(ages, c(i, i+4)) }
ages <- c(ages, c(81,100)) # The last age range is 81 plus
for (i in seq(1,length(ages)-1, 2)) # Each pair of ages in the age ranges
{ # Construct the column using apply, Paste here concatenates the string for the column name
  gvd2015_keep_cali[paste("age", ages[i],ages[i+1], sep="_")] <- apply(gvd2015_keep_cali, 1, function(row) {
    cnt <- 0 # Count of ages in current range starts at 0
    for (x in row$matches) # Iterate through the matches we parsed out
    { if (!grepl(":", x, fixed = TRUE)) # If the match does not contain a : then it is the match group rather than the match
      { age <- as.integer(x) # so convert age as string to integer
        if (age >= ages[i] && age <= ages[i+1]) # and check if age falls within age range for the current column
        { cnt <- cnt + 1 }}}
    cnt })}
```

We will split gender into M | F | U (U for unknown)

```{r}
gvd2015_keep_cali$participant_gender <-  gvd2015_keep_cali$participant_gender %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_gender_unique_check <- as.character(gvd2015_keep_cali[,c("participant_gender")])
ul <- unlist(strsplit(participant_gender_unique_check, " "))
unique(ul)
```

```{r}
gvd2015_keep_cali$male_count <- str_count(gvd2015_keep_cali$participant_gender, pattern = "Male") # Stringr
gvd2015_keep_cali$female_count <- str_count(gvd2015_keep_cali$participant_gender, pattern = "Female") # Stringr
```

```{r}
colnames(gvd2015_keep_cali) # Check column names haven't changed!!!
gvd2015_keep_cali[1:6, c(13,34,35)]
```

#### Participant Age Group Column
```{r}
gvd2015_keep_cali$participant_age_group <-  gvd2015_keep_cali$participant_age_group %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_age_group_unique_check <- as.character(gvd2015_keep_cali[,c("participant_age_group")])
ul <- unlist(strsplit(participant_age_group_unique_check, " "))
unique(ul)
```

DROP THIS

```{r}
colnames(gvd2015_keep_cali)
```

```{r}
gvd2015_keep_cali$participant_status <-  gvd2015_keep_cali$participant_status %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_status_unique_check <- as.character(gvd2015_keep_cali[,c("participant_status")])
ul <- unlist(strsplit(participant_status_unique_check, " "))
unique(ul)
```

```{r}
InjuredUnharmedArrested <- grepl("InjuredUnharmedArrested", gvd2015_keep_cali$participant_status)
summary(InjuredUnharmedArrested)
gvd2015_keep_cali[InjuredUnharmedArrested,]
```

```{r}
filter(gvd2015, incident_id == 273316) # http://www.gunviolencearchive.org/incident/273316
```

CHANGE TO INJURED ARRESTED

Mention URL stem ...

```{r}
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "InjuredUnharmedArrested", "InjuredArrested")
```



```{r}
KilledArrested <- grepl("KilledArrested", gvd2015_keep_cali$participant_status)
summary(KilledArrested)
gvd2015_keep_cali[KilledArrested,]
```

It was determined that "KilledArrested" means they were arrested but lated died from injuries sustained. Change to InjuredArrestedDied.

```{r}
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledArrested", "InjuredArrestedDied")
```


```{r}
KilledUnharmed <- grepl("KilledUnharmed", gvd2015_keep_cali$participant_status)
summary(KilledUnharmed)
gvd2015_keep_cali[KilledUnharmed,]
```

Incident 318320 was checked online, should be killed, injured, killed.
Incident 440261 should be killed killed

```{r}
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledUnharmed", "Killed")
```

```{r}
participant_status_unique_check <- as.character(gvd2015_keep_cali[,c("participant_status")])
ul <- unlist(strsplit(participant_status_unique_check, " "))
unique(ul)
```

```{r}
# Check WORD COUNT
gvd2015_keep_cali$Killed_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Killed") # Stringr
gvd2015_keep_cali$Injured_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Injured") # Stringr
gvd2015_keep_cali$UnharmedArrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "UnharmedArrested") # Stringr (with anchor)
gvd2015_keep_cali$InjuredArrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "InjuredArrested") # Stringr (with anchor)
gvd2015_keep_cali$Arrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Arrested") # Stringr (with anchor)
gvd2015_keep_cali$Unharmed_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Unharmed") # Stringr (with anchor)
gvd2015_keep_cali$InjuredArrestedDied_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "InjuredArrestedDied") # Stringr
```

```{r}
# Check if this is needed???
KilledUnharmedArrested <- grepl("KilledUnharmedArrested", gvd2015_keep_cali$participant_status)
summary(KilledUnharmedArrested)
gvd2015_keep_cali[KilledUnharmedArrested,]
```

Should be Killed and UnharmedArrested

```{r}
colnames(gvd2015_keep_cali)
```

```{r}
gvd2015_pt2 <- select(gvd2015_keep_cali, incident_id, city_or_county, n_killed, n_injured, congressional_district, latitude, longitude, n_guns_involved, participant_type, state_house_district, state_senate_district, age_0_15, age_16_20, age_21_25, age_26_30, age_31_35, age_36_40, age_41_45, age_46_50, age_51_55, age_56_60, age_61_65, age_66_70, age_71_75, age_76_80, age_81_100, male_count, female_count, Killed_count, Injured_count, UnharmedArrested_count, InjuredArrested_count, Arrested_count, Unharmed_count, InjuredArrestedDied_count) ##### Unfortuanetly ORDER has not been preserved...
```




### ACS

```{r Column Names, message=FALSE, warning=FALSE, paged.print=TRUE}
dim(acs2015)
```

```{r Head of acs2015Pt1, message=FALSE, warning=FALSE, paged.print=TRUE}
#head(acs2015, n=8)
```

```{r colnames acs2015, message=FALSE, warning=FALSE, paged.print=TRUE}
colnames(acs2015) # Checking the column names
```

```{r}
sapply(acs2015, class) # Checking the data types for 'acs2015'
```




##### Comment - drop columns first then join then clean... OR clean then join...

For the following analysis, nly interested in economic indicators

```{r}
acs2015_keep <- select(acs2015, State, County, TotalPop, Income, IncomePerCap, Poverty, Unemployment) #Dplyr
```

Checking the data types

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
sapply(acs2015_keep, class)
```

```{r}
summary(acs2015_keep)
```

```{r}
aggregate(County ~ State, acs2015_keep, function(x) length(unique(x)))
```

```{r}
acs2015_keep_cali <- filter(acs2015_keep, State == "California")
```

```{r}
aggregate(County ~ State, acs2015_keep_cali, function(x) length(unique(x)))
```

```{r}
aggregate(city_or_county ~ city_or_county, gvd2015_pt2, function(x) length(unique(x)))
```


### Merge ... 

```{r}
colnames(gvd2015_pt2)
gvd2015_pt2 = rename(gvd2015_pt2, County = city_or_county)
colnames(gvd2015_pt2)
```

```{r}
join <- inner_join(gvd2015_pt2, acs2015_keep_cali, by = "County") # Disable warnings...
```

#### How could CENSUS data be missing???? Justifiably there may not be gun incidents in every country, but there IS more info that gun violence!!!!




















 

```{r}
# participant_gender string count -> Then find that string!!!
#gvd2015_keep_cali_age_gender$participant_gender <- as.character(gvd2015_keep_cali_age_gender$participant_gender)
#gvd2015_keep_cali_age_gender$genderChar <- nchar(gvd2015_keep_cali_age_gender$participant_gender, 
                                             #    type = "chars", allowNA = FALSE, keepNA = NA)
```

```{r}
#max(gvd2015_keep_cali_age_gender$genderChar)
#which.max(gvd2015_keep_cali_age_gender$genderChar)
```

```{r}
#gvd2015_keep_cali_age_gender[2977,]
```

http://stringr.tidyverse.org/articles/regular-expressions.html

```{r}
#gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "(\\d+)", "")
```


participant_status_only <- as.data.frame(gvd2015_keep_cali[,c("participant_status")])
colnames(participant_status_only) <- c("col1")
participant_status_only <- unique(participant_status_only) # 155 obs left
which.max(participant_status_only$col1) # [1] 52
participant_status_only[52,]
participant_status_only$col1 <- trimws(participant_status_only)
participant_status_only <- separate(participant_status_only, col1, into = c('1','2','3','4','5','6','7','8','9','10'), sep = " ")



column1 %>% drop_na()

trimws()

separate(rate, into = c("cases", "population"), sep = "/")

listCheckCollapse <- paste(listCheck, collapse = '') # class(listCheckCollapse) = "character"
listCheckCollapse <- as.vector(listCheckCollapse)
unique(listCheckCollapse) # 
listCheckCollapse
unique(unlist(listCheckCollapse))


test <- separate(listCheck, participant_status, into = c(1,2,3,4,5,6,7,8,9,10), sep = " ")
test <- separate(listCheck, participant_status, into = c(1,2,3,4,5,6,7,8,9,10), sep = " ")


```{r}
#remove_all_characters <- function(dataset, col_name) {
#  dataset$col_name <- NA
#  dataset$col_name <-  dataset$col_name %>% 
#  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
#  str_replace_all("::", "") %>%
#  str_replace_all("[:digit:]", " ") %>%
#  str_replace_all("\\||", "")
#  return(dataset)
#}
```

```{r}
#try <- remove_all_characters(gvd2015_keep_cali, participant_gender)
```



```{r}
# participant_gender string count -> Then find that string!!!
#gvd2015_keep_cali_age_gender$participant_gender <- as.character(gvd2015_keep_cali_age_gender$participant_gender)
#gvd2015_keep_cali_age_gender$genderChar <- nchar(gvd2015_keep_cali_age_gender$participant_gender, 
                                               #  type = "chars", allowNA = FALSE, keepNA = NA)
```

```{r}
#max(gvd2015_keep_cali_age_gender$genderChar)
#which.max(gvd2015_keep_cali_age_gender$genderChar)
```

```{r}
#gvd2015_keep_cali_age_gender[2977,]
```


```{r}
#acs2015_keep$SC_concat <- do.call(str_c, c(acs2015_keep[c("State", "County")], sep = "_")) 
```

gvd$date_year <- format.Date(gvd$date, format="%Y") # Converting YYYY-DD-MM to


gvd2015 <- gvd %>% filter(between(date, "2015-01-01", "2015-12-13"))
  
gvd$date >= as.Date("2015-01-01") & gvd$date >= as.Date("2015-12-31")

year2015_boolean <- as.logical(year(gvd$date) == "2015")

year2015_boolean <- as.logical(format.Date(gvd$date, format="%Y"))
summary(year2015_boolean)
year2015_boolean

```{r}
#str(gvd$date_year)
```

```{r}
#gvd$date_year <- format.Date(gvd$date, format="%Y") # Converting YYYY-DD-MM to
#gvd$date_year <- filter(gvd, date_year == "2015") # Filtering Gun Violence Data for 2015 only
```



