---
#title: "US Gun Violence and Census Data from California 2015"
#author: "Alistair Grevis-James (s3644119) and Chris Kaias (sXXXXXX)"
#subtitle: Assignment 3, MATH2349, Semester 1, 2018 
#output:
#  html_notebook:
#      toc: true
#      toc_float: true
---

<style>
body {
text-align: justify}
</style>

```{r, include = FALSE}
options(knitr.table.format = "html")
```

### Executive Summary

The required libraries were imported into the RStudio interactive development environment, followed by the .csv files “gun-violence-data” and “American Census Data 2015” as `gvd` and `acs2015`, respectively [^1][^2]. These data were then checked for tidiness (`gvd` untidy, `acs2015` tidy), dimensions (`gvd` 239677 obs. of 29 var., `acs2015` 3220 obs. of 37 var.), NA count and attribute classes (see summary tables, below). 
The `gvd` dataset failed to conform to the “tidy data” principals. Most notably, the attributes prefixed “participant_” contained multiple values per cell, delimited by a combination of [:digit:], :: and/or ||. In order to be computationally efficient, it was decided to first subset `gvd` before cleaning the aforementioned “participant_” attributes.  As the `acs2015` dataset contained values from only the 2015 American (US) Census, the first subset involved filtering values of the year 2015 from the `gvd` dataset. 
This was achieved using the `as.logical()` and `year()` functions, creating and applying a boolean vector to `gvd` to create the dataset `gvd2015`. This methodology enabled us to retain the native date format YYYY-MM-DD. The `gvd2015` was then filtered for values which satisfied the condition `state == “California”`, giving `gvd2015_keep_cali`. A custom function was written and employed to scour through the `participant_age` column of `gvd2015`. This resulted in capturing age counts of incident participants by binning them into approximately 5-years bin intervals. The `participant, _gender, _status, _type` columns were then also all stripped of string prefix values and binned. The resultant `killed_count` and `injured_count` matched perfectly the original `n_killed` and `n_injured` attributes, indicating an extremely precise processing methodology. It was determined that the `incident_url` column of `gvd2015_keep_cali` could be dropped, as the URL prefix in every case matched the unique incident ID.
The `acs2015` dataset was then filtered for values from the state of California.


### Required packages 

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
#The following packages were used whilst completing the report (with annotation)
library(readr) # Used to import .csv files
library(dplyr) # Section Y
library(stringr) # Section Z
library(knitr) # Section X
library(tidyr)
library(lubridate)
library(kableExtra)
library(outliers)
library(ggplot2)
#xlsx #readxl #foreign #gdata #rvest #dplyr  #deductive #validate #Hmisc 
#MVN #infotheo #MASS #caret #MLR #ggplot #base R functions
```

### Importing the Data

The first dataset, 'Gun Violence Data - A Comprehensive Record of Over 260k US Gun Violence Incidents from 2013-2018', was compiled by James Ko and made available on kaggle.com[^1]. These data contain all recorded gun violence incidents in the US between January 2013 and March 2018, inclusive.
<br>
The second dataset, 'US Census Demographic Data Demographic and Economic Data for Tracts and Counties', was collected by the US Census Bureau and made available on kaggle.com[^2]. These data contain all census data from the 2015 federal US census. 
<br>
The `readr` function was used to import the data as shown below.

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
gvdZip <- "gun-violence-data.csv.zip"
outDir<-"/Users/alistairgj/Documents/GitHub" # Change output directory for unzipped .csv file as needed
unzip(gvdZip, exdir=outDir)
setwd("/Users/alistairgj/Documents/GitHub")
gvd <- read.csv("gun-violence-data.csv") # Importing Gun Violence Data
setwd("/Users/alistairgj/Documents/GitHub/MATH2349_Assignment3V2")
acs2015 <- read.csv("acs2015_county_data.csv") # Importing American (US) Census Data 2015
```

### Checking the Data

Two custom output table was created to check `gvd` & `acs2015`. These results are summarized below.

 Feature 		| gvd 					| acs2015
------------- 	      | ------------------------------ | -----------
Observations   | 239677 | 3220
Attributes   | 29 |  37
Tidy | No | Yes
Format | Wide | Wide 
Unique Ref   | Yes - IncidentID | Yes - CensusID 
Incorrect Class  | Attribute 1, 5, 8:14, 16, 19 |  Attribute 1
Contains NA  | Attribute 11, 15, 17, 18, 28, 29 | Attribute 14, 15, 19

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(AttNo = c(1:29), Attribute = names(gvd), Class = sapply(gvd, class),
           gvd_incident486623 = sapply(gvd[84,], function(x) paste0(head(x),  collapse = ", ")),
           NA_Count = sapply(gvd, function(y) sum(length(which(is.na(y))))),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% scroll_box(height = "200px")
```

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(AttNo = c(1:37), Attribute = names(acs2015), Class = sapply(acs2015, class),
           acs2015 = sapply(acs2015, function(x) paste(head(x),  collapse = ", ")),
           NA_Count = sapply(acs2015, function(y) sum(length(which(is.na(y))))),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                                                           font_size = 11) %>% scroll_box(height = "200px")
```

The `length()` & `unique()` functions were used to determine if `gvd$incident_id` and `acs2015$CensusId` contained only unique values. In each case the number of values was equal to the number of observations. Thus `incident_id` and `CensusId` may be used as observation identifiers if required.

```{r}
length(unique(gvd$incident_id)) # Checking the number of unique values for incident_id
```

```{r}
length(unique(acs2015$CensusId)) # Checking the number of unique values for CensusId
```

Based on the output of the above `gvd` summary table (`data.frame(...`), the `participant_age` and `participant_gender` were explored to better understand their values. It was noted that these cells store multiple, discrete instances of data. The [:digit:] at the start of each value is presumably used to track the values accross from one `participant_` attribute to the next. It is also worth noting missing values. The first observation below contains only one age value, with four participant values. These attributes will be address later in the analysis.

```{r}
gvd %>% select(participant_age, participant_gender) %>% head(n=8) %>% 
  kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) 
```

### Filtering GVD for Year 2015

The Gun Violence Dataset contains data from 2013 till 2015, while the American Census Data is from 2015. We will therefore create a data subset of the Gun Violence Data from the year 2015 only (as we will later be joining these data). The attribute `gvd$date` was converted to date format (YYYY-MM-DD) using `as.Date` and checked with `str()`, below. A Boolean vector was then created for 2015-MM-DD values. This was applied as a mask to `gvd`, and extracted as the subset `gvd2015`, which was then checked.

```{r}
gvd$date <- as.Date(gvd$date) # Converting the gvd$date column and applying to the gvd dataset
str(gvd$date) # Checking the conversion of gvd$date
```

```{r}
year2015_boolean <- as.logical(year(gvd$date) == "2015") #Creating a year 2015 true/false vector
summary(year2015_boolean) # Checking the vector results
```

```{r}
gvd2015 <- gvd[year2015_boolean,] # Applying the True/False Vector to filter gvd
```

```{r}
summary(gvd2015$date) # Confirming only 2015 data is present
```

The DS `gvd2015` is now prepared. As an aside, the first methodology used for extracting YYYY = 2015 value from `gvd$date` relied on conversion of `gvd$date` to contain year only (`gvd$date <- format.Date(gvd$date, format="%Y")`), this was then filtered for values of 2015 (`gvd$date_year <- filter(gvd, date_year == "2015")`). This methodolgy was abandoned, as it unnesscesarily stripped out the month and day data from `gvd$date`. The applied method is no more complex but retains more information.


### Filtering Gun Violence Data 2015 for values from the state of California

The values of `gvd2015$state` were aggregated for comparison (below). The output enables us to determine the number of `city_or_county` in each State, as defined during the composition of the Gun Violence Dataset. For the subsequent analysis, the State of California was selected. This was based on an assumption it would have a large range of discrete and continuous values in the corresponding Census dataset (for example extremely poor and rich socioeconomic areas). Also note that the methodology outlined below could be applied to any of the states in the `gvd` dataset, so California will act as a proof of concept.

```{r}
aggregate(city_or_county ~ state, gvd2015, function(x) length(unique(x))) %>% 
  kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% 
  row_spec(5, bold = T, background = "lightskyblue") %>%
  scroll_box(height = "200px")
```

```{r}
gvd2015_keep_cali <- filter(gvd2015, state == "California") # Creating the new dataset
```

```{r}
aggregate(city_or_county ~ state, gvd2015_keep_cali, function(x) length(unique(x))) %>% 
  kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% 
  row_spec(1, bold = T, background = "lightskyblue") # Check the new dataset
```

### Exploring gvd2015_keep_cali

During data exploration it was determined the following attributes will add no futher value to the analysis and will be dropped.

 Attribute | Explanation / Justification
------------- | ------------------------------------------------- 
`address` | This level of extremely fine detail is not required
`source_url` | As this can be found online via `incident_url` 
`incident_url_fields_missing` | As this is a boolean attribute which can be checked with filtering
`incident_characteristics` | As these strings are highly unstructured 
`location_description` | These strings are highly unstructured
`notes` | Contains no information of value for this analysis
`participant_age_group` | We can determine this from `participant_age` 
`participant_name` | This level of extremely fine detail is not required
`sources` | We have source information from `incident_url`

The desired columns were selected back into `gvd2015_keep_cali`, which was then checked in the table below. The attributes `congressional_district`, `state_house_district` & `state_sentate_district` were converted into factors and the dimensions of the new `gvd2015_keep_cali` were checked.

```{r}
gvd2015_keep_cali <- select(gvd2015_keep_cali, incident_id, date, state, city_or_county, n_killed, n_injured, incident_url, congressional_district, latitude, longitude, n_guns_involved, participant_age, participant_gender, participant_status, participant_type, state_house_district, state_senate_district)
```

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(Attribute = names(gvd2015_keep_cali), Class = sapply(gvd2015_keep_cali, class),
           gvd_incident274168 = sapply(gvd2015_keep_cali[53,], function(x) paste0(head(x),  collapse = ", ")),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                                                           font_size = 11) %>% scroll_box(height = "200px")
```

```{r}
gvd2015_keep_cali$congressional_district <- as.factor(gvd2015_keep_cali$congressional_district)
gvd2015_keep_cali$state_house_district <- as.factor(gvd2015_keep_cali$state_house_district)
gvd2015_keep_cali$state_senate_district <- as.factor(gvd2015_keep_cali$state_senate_district)
```

```{r}
dim(gvd2015_keep_cali)
```

#### Incident_URL

The attribute `incident_URL` will now be compared with `incident_ID` (as we may be able to drop `incident_URL`)

```{r}
gvd2015_keep_cali$incident_url <-  gvd2015_keep_cali$incident_url %>% 
  str_replace_all("http://www.gunviolencearchive.org/incident/", "") # 
gvd2015_keep_cali$incident_url <- as.integer(gvd2015_keep_cali$incident_url)
```

```{r}
# Create two vectors
a <- as.vector(gvd2015_keep_cali$incident_id)
b <- as.vector(gvd2015_keep_cali$incident_url)
a[!(a %in% b)]
```

- Drop incident_url and will use incident_id with URL stem.

#### Exploring the GVD 2015 California Data Subset

* Text text text
* The participant age and gender attributes may contain ... (good data for analytics)
* In both attibutes, multiple values are contained. 
* Not tidy, multiple values in one column

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
gvd2015_keep_cali[1:5, c(11:15)] %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```

##### n_gun_involved

```{r}
table(sapply(gvd2015_keep_cali$n_guns_involved, function(y) sum(length(which(is.na(y)))))) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) 
```

- Almost 50% NA - drop
- Futher analysis = machine learning

##### participant_age

- Current form
- Discuss

```{r}
gvd2015_keep_cali$matches <- str_match_all(gvd2015_keep_cali$participant_age, "\\d+::(\\d+)")
ages = c(c(0,15)) # First age range is under 15s
for (i in seq(16, 76, 5)) # The other age ranges are every 5 years 
{
  ages <- c(ages, c(i, i+4))
}
ages <- c(ages, c(81,100)) # The last age range is 81 plus
for (i in seq(1,length(ages)-1, 2)) # Each pair of ages in the age ranges
{
  # Construct the column using apply
  # Paste here concatenates the string for the column name
  gvd2015_keep_cali[paste("age", ages[i],ages[i+1], sep="_")] <- apply(gvd2015_keep_cali, 1, function(row) {
    cnt <- 0 # Count of ages in current range starts at 0
    for (x in row$matches) # Iterate through the matches we parsed out
    { if (!grepl(":", x, fixed = TRUE)) # If the match does not contain a : then it is the match group rather than the match
      { age <- as.integer(x) # so convert age as string to integer
          if (age >= ages[i] && age <= ages[i+1]) # and check if age falls within age range for the current column 
        { cnt <- cnt + 1 }
      }
    }
    as.integer(cnt)
  })
}
```

##### participant_gender

The characters `,[:space:]`, `::`, `[:digit:]` & `||` were stripped from the `participant_gender` as seen below. A new char value was subsequently created, `participant_gender_unique_check`. This value was then checked for unique characters, which yielded only blank (""), Female and Male. Two `str_count` functions were then used to generate count columns for gender, these were named `male_count` and `female_count`. The column `participant_gender` will be later dropped.

```{r}
gvd2015_keep_cali$participant_gender <-  gvd2015_keep_cali$participant_gender %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_gender_unique_check <- as.character(gvd2015_keep_cali[,c("participant_gender")])
ul <- unlist(strsplit(participant_gender_unique_check, " "))
unique(ul)
```

```{r}
gvd2015_keep_cali$male_count <- str_count(gvd2015_keep_cali$participant_gender, pattern = "Male") # Stringr
gvd2015_keep_cali$female_count <- str_count(gvd2015_keep_cali$participant_gender, pattern = "Female") # Stringr
```

##### Participant_status

The characters `,[:space:]`, `::`, `[:digit:]` & `||` were stripped from the `participant_status` in the same manner as for `participant_gender`. The same methodology was then used to generate counts (`unlist`, `strsplit`, `unique`) for the `participant_status` values, giving the following;

```{r, include = FALSE}
gvd2015_keep_cali$participant_status <-  gvd2015_keep_cali$participant_status %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_status_unique_check <- as.character(gvd2015_keep_cali[,c("participant_status")])
ul <- unlist(strsplit(participant_status_unique_check, " "))
unique(ul)
```

Several anomalous values which contain mutually exclusive states of being, were identified as; "InjuredUnharmedArrested", "KilledInjured", "KilledArrested", "KilledUnharmed", "InjuredUnharmed" and "KilledUnharmedArrested". The `grepl` function was then used to create a boolean vector for in each case to search for the source of the value in `gvd2015_keep_cali`. Once the `incident_id` was identified, the URL "http://www.gunviolencearchive.org/incident/`incident_id`" was checked. Each instance was then rationalized and given a more appropriate value using the `str_replace_all` function. 

```{r}
InjuredUnharmedArrested <- grepl("InjuredUnharmedArrested", gvd2015_keep_cali$participant_status)
gvd2015_keep_cali[InjuredUnharmedArrested,c(1,14)]
```

The values were all checked as described above and updated as follows;

 OriginalValue | Explanation / Justification | UpdatedValue
------------- | ------------------------------------------------- | -----------
`InjuredUnharmedArrested` | This value was created as the reporting of an incident is updated | `InjuredArrested`
`KilledInjured` | This value was created when a person who was initially injured later dies | `Killed`
`KilledArrested` | This value was created when a person who was mortally wounded, arrested and later dies | `Killed`
`KilledUnharmed` | This value was created as the reporting of an incident is updated | `Killed`
`InjuredUnharmed` | This value was created as the reporting of an incident is updated | `Injured`
`KilledUnharmedArrested` | This value was created as the reporting of an incident is updated | `Killed`

```{r}
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "InjuredUnharmedArrested", "InjuredArrested")
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledInjured", "Killed")
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledArrested", "Killed")
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledUnharmed", "Killed")
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "InjuredUnharmed", "Injured")
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledUnharmedArrested", "Killed")
```

```{r, include=FALSE}
gvd2015_keep_cali$participant_status <- str_replace_all(gvd2015_keep_cali$participant_status, "KilledArrested", "Killed")
```

```{r}
participant_status_unique_check <- as.character(gvd2015_keep_cali[,c("participant_status")])
ul <- unlist(strsplit(participant_status_unique_check, " "))
unique(ul)
```

```{r}
gvd2015_keep_cali$injured_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Injured")
gvd2015_keep_cali$unharmed_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Unharmed")
gvd2015_keep_cali$killed_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Killed")
gvd2015_keep_cali$unharmedArrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "UnharmedArrested")
gvd2015_keep_cali$arrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "Arrested")
gvd2015_keep_cali$injuredArrested_count <- str_count(gvd2015_keep_cali$participant_status, pattern = "InjuredArrested") 
```


##### Participant_Type

The characters `,[:space:]`, `::`, `[:digit:]` & `||` were stripped from the `participant_type` in the same manner as for `participant_gender`. The same methodology was then used to generate counts (`unlist`, `strsplit`, `unique`) for the `participant_type` values, giving the following;

```{r, include = FALSE}
gvd2015_keep_cali$participant_type <-  gvd2015_keep_cali$participant_type %>% 
  str_replace_all(",[:space:]", "") %>% #Remove comma & whitespace
  str_replace_all("::", "") %>%
  str_replace_all("[:digit:]", " ") %>%
  str_replace_all("\\||", "")
```

```{r}
participant_type_unique_check <- as.character(gvd2015_keep_cali[,c("participant_type")])
ul <- unlist(strsplit(participant_type_unique_check, " "))
unique(ul)
```

```{r}
gvd2015_keep_cali$victim_count <- str_count(gvd2015_keep_cali$participant_type, pattern = "Victim")
gvd2015_keep_cali$subjectSuspect_count <- str_count(gvd2015_keep_cali$participant_type, pattern = "Subject-Suspect")
```

#### The Date Column ???

```{r}
colnames(gvd2015_keep_cali)
```

```{r}
# Checking
a <- as.vector(gvd2015_keep_cali$n_killed)
b <- as.vector(gvd2015_keep_cali$killed_count)
a[!(a %in% b)]
```

```{r}
# Checking
a <- as.vector(gvd2015_keep_cali$n_injured)
b <- as.vector(gvd2015_keep_cali$injured_count)
a[!(a %in% b)]
```

### 

```{r}
gvd2015_pt2 <- select(gvd2015_keep_cali, incident_id, date, state, city_or_county, congressional_district, state_house_district, state_senate_district, latitude, longitude, age_0_15, age_16_20, age_21_25, age_26_30, age_31_35, age_36_40, age_41_45, age_46_50, age_51_55, age_56_60, age_61_65, age_66_70, age_71_75, age_76_80, age_81_100, male_count, female_count, injured_count, unharmed_count, killed_count, unharmedArrested_count, arrested_count, injuredArrested_count, victim_count, subjectSuspect_count)
```

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(Attribute = names(gvd2015_pt2), Class = sapply(gvd2015_pt2, class),
           gvd2015_keep_cali = sapply(gvd2015_pt2, function(x) paste(head(x),  collapse = ", ")),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```

### ACS

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(Attribute = names(acs2015), Class = sapply(acs2015, class),
           acs2015 = sapply(acs2015, function(x) paste(head(x),  collapse = ", ")),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% scroll_box(height = "200px")
```

##### Comment - drop columns first then join then clean... OR clean then join...

For the following analysis, only interested in economic indicators and population

```{r}
acs2015_keep <- select(acs2015, State, County, TotalPop, Men, Women, Income, IncomePerCap, Poverty, Unemployment)
```

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(Attribute = names(acs2015_keep), Class = sapply(acs2015_keep, class),
           acs2015 = sapply(acs2015_keep, function(x) paste(head(x),  collapse = ", ")),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% scroll_box(height = "200px")
```

```{r}
aggregate(County ~ State, acs2015_keep, function(x) length(unique(x))) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) %>% scroll_box(height = "200px")
```

```{r}
acs2015_keep_cali <- filter(acs2015_keep, State == "California")
```

```{r}
aggregate(County ~ State, acs2015_keep_cali, function(x) length(unique(x))) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```

```{r}
aggregate(city_or_county ~ city_or_county, gvd2015_pt2, function(x) length(unique(x))) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```


### Merge ... 

```{r, warning=FALSE}
gvd2015_pt2 <- rename(gvd2015_pt2, County = city_or_county)
gvd_acs_join <- inner_join(gvd2015_pt2, acs2015_keep_cali, by = "County")
gvd_acs_join <- subset(gvd_acs_join, select = -c(State, state))
```


### Merge ... 

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
data.frame(Attribute = names(gvd_acs_join), Class = sapply(gvd_acs_join, class),
           gvd_acs_join = sapply(gvd_acs_join, function(x) paste0(head(x),  collapse = ", ")),
           NA_count = sapply(gvd_acs_join, function(y) sum(length(which(is.na(y))))),
           row.names = NULL) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11) #%>% scroll_box(height = "200px")
```

#### NaN Values

```{r}
gvd_acs_join %>% filter(is.na(congressional_district)) %>% 
  select(County, congressional_district, state_house_district, state_senate_district, latitude, longitude) %>%
  kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```

The aggregate function was used to determine the number of congressional districts in each county.

```{r echo=FALSE}
gvd_acs_join_fresno <- gvd_acs_join %>% filter(County == "Fresno")
one <- aggregate(congressional_district ~ County, gvd_acs_join_fresno, function(x) length(unique(x))) 
gvd_acs_join_sanBernardino <- gvd_acs_join %>% filter(County == "San Bernardino")
two <- aggregate(congressional_district ~ County, gvd_acs_join_sanBernardino, function(x) length(unique(x)))
gvd_acs_join_losAngeles <- gvd_acs_join %>% filter(County == "Los Angeles")
three <- aggregate(congressional_district ~ County, gvd_acs_join_losAngeles, function(x) length(unique(x))) 
gvd_acs_join_sacramento <- gvd_acs_join %>% filter(County == "Sacramento")
four <- aggregate(congressional_district ~ County, gvd_acs_join_sacramento, function(x) length(unique(x)))
c <- union_all(one, two)
d <- union_all(three, four)
union_all(c, d) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```

```{r}
gvd_acs_join <- na.omit(gvd_acs_join)
```

```{r echo=FALSE}
gvd_acs_join %>% filter(is.na(congressional_district)) %>% 
  select(County, congressional_district, state_house_district, state_senate_district, latitude, longitude) %>%
  kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  font_size = 11)
```


```{r}
hist(gvd_acs_join$Income)
```

```{r}
library(forecast)
boxcox_income <- BoxCox(gvd_acs_join$Income,lambda = "auto")

hist(boxcox_income)

recip <- 1/gvd_acs_join$Income
hist(recip)
```


```{r}
hist(gvd_acs_join$male_count)
```

```{r}
x3recip <- 1/gvd_acs_join$male_count
hist(x3recip)
```

```{r}
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
gvd_acs_join$latitude_norm <- range01(gvd_acs_join$latitude)
gvd_acs_join$longitude_norm <- range01(gvd_acs_join$longitude)
```

```{r}
hist(gvd_acs_join$longitude)
```

```{r}
hist(gvd_acs_join$latitude_norm)
```

```{r}
hist(gvd_acs_join$longitude_norm)
```


```{r}
boxplot(gvd_acs_join$killed_count ~ gvd_acs_join$MONTH, main="Diamond carat by cut", ylab = "Carat", xlab = "Cut")
```


```{r}
# Log 10 apply
gvd_acs_join %>% plot(Unemployment ~ Income, data = ., ylab="Unemployment", xlab="Income", main="Carat by Depth")
```

```{r}
# GGplot2 - bubbleplot
gvd_acs_join %>% plot(latitude_norm ~ longitude_norm, data = ., ylab="Unemployment", xlab="Income", main="Carat by Depth")
```

```{r}
gvd_acs_join$IncomePerCap %>%  boxplot(main="Box Plot of Diamond Carat", ylab="Carat", col = "grey")
```


```{r}
# Checking the summary stats
summary(gvd_acs_join$IncomePerCap)
```

```{r}
# Checking the z score summary stats
z.scores <- gvd_acs_join$IncomePerCap %>% scores(type = "z")
z.scores %>% summary()
```

```{r}
# Checking the z score values
gvd_acs_join$IncomePerCap[which(abs(z.scores) > 3)]
```

```{r}
# Imputing the outliers
surveys_combined$hindfoot_length[ which( abs(z.scores) >3 )] <- mean(surveys_combined$hindfoot_length, 
                                                                     na.rm = TRUE)
```

```{r}
# Checking the results
surveys_combined$hindfoot_length %>%  boxplot(main="Box Plot of Hindfoot Length", 
                                              ylab="Hindfoot Length", col = "grey")
```

Outliers were removed by imputation. The method of outlier removal in the case of these data does not require a large amount of consideration. This is because the hindfoot length is being taken over a huge range of different species, so this statistical evaluation has minimal applicable value. Additionally the outliers reprented only four values out of over 30,000, so any method do deal with the outliers will have minimal effect.

```{r}
gvd_acs_join$Unemployment %>%  boxplot(main="Box Plot of Diamond Carat", ylab="Carat", col = "grey")
```

```{r}
gvd_acs_join$killed_count %>%  boxplot(main="Box Plot of Diamond Carat", ylab="Carat", col = "grey")
```

```{r}
# Checking the summary stats
summary(gvd_acs_join$killed_count)
```

```{r}
# Checking the z score summary stats
z.scores <- gvd_acs_join$killed_count %>% scores(type = "z")
z.scores %>% summary()
```

```{r}
# Checking the z score values
gvd_acs_join$killed_count[which(abs(z.scores) > 3)]
```

```{r}
# Imputing the outliers
killed_count_clean <- gvd_acs_join$killed_count[ - which(abs(z.scores) > 3)]
```

```{r}
killed_count_clean %>%  boxplot(main="Box Plot of Diamond Carat", ylab="Carat", col = "grey")
```




### References (see last page)
[^1]: https://www.kaggle.com/jameslko/gun-violence-data/data
[^2]: https://www.kaggle.com/zimeiyang/2015-us-census-demographic-data/data






