

## Required packages 


Provide the packages required to reproduce the report. Make sure you fulfilled the minimum requirement #10.

```{r}

```


## Executive Summary 


In your own words, provide a brief summary of the preprocessing. Explain the steps that you have taken to preprocess your data. Write this section last after you have performed all data preprocessing. (Word count Max: 300 words)


## Data 

A clear description of data sets, their sources, and variable descriptions should be provided. In this section, you must also provide the R codes with outputs (head of data sets) that you used to import/read/scrape the data set. You need to fulfil the minimum requirement #1 and merge at least two data sets to create the one you are going to work on. In addition to the R codes and outputs, you need to explain the steps that you have taken.


```{r}

```

## Understand 

Summarise the types of variables and data structures, check the attributes in the data. In addition to the R codes and outputs, explain briefly the steps that you have taken. In this section, show that you have fulfilled minimum requirements 2-4.

```{r}

```


##	Tidy & Manipulate Data I 

Check if the data conforms the tidy data principles. If your data is not in a tidy format, reshape your data into a tidy format (minimum requirement #5). In addition to the R codes and outputs, explain everything that you do in this step.


```{r}

```

##	Tidy & Manipulate Data II 

Create/mutate at least one variable from the existing variables (minimum requirement #6). In addition to the R codes and outputs, explain everything that you do in this step..

```{r}

```


##	Scan I 

Scan the data for missing values, inconsistencies and obvious errors. In this step, you should fulfil the minimum requirement #7. In addition to the R codes and outputs, explain how you dealt with these values.

```{r}

```


##	Scan II

Scan the numeric data for outliers. In this step, you should fulfil the minimum requirement #8. In addition to the R codes and outputs, explain how you dealt with these values.

```{r}

```


##	Transform 

Apply an appropriate transformation for at least one of the variables. In addition to the R codes and outputs, explain everything that you do in this step. In this step, you should fulfil the minimum requirement #9.

```{r}

```


NOTE: Follow the order outlined above in the report. Make sure your code is visible (within the margin of the page). Do not use View() to show your data instead give headers (using head() )


Any further or optional pre-processing tasks can be added to the template using an additional section in the R Markdown file. Please also provide the R codes, outputs and brief explanations on why and how you applied these tasks on the data.

<br>
<br>
